---
title: "Can we tell the difference between text, non-text and random data?"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
---

# Motivation

I read the *Why 83?* document and wanted to try a different statistical analysis.

# Sub-questions
## Are messages correlated to the location they spawn in?

The underlying assumption in this section is that *eye_state* is a categorical 
variable. We'll be looking at the $\chi^2$ and $p_i - p_j$ statistics.

For the purposes of this section *eye messages* are the equivalent of a 5-choice 
response to a poll.
\newline
```{r}
options(warn=FALSE)
suppressMessages(library(ggplot2))

messages = read.csv('./single_message.csv')
ggplot(messages, aes(y=factor(eye_state), fill=factor(location))) + 
  geom_bar(position='fill') +
  scale_fill_brewer(palette='PRGn') +
  labs(title='Relative proportion of occurrences of eye states', 
       subtitle='Aggregated by location',
       y='Eye state', x='Relative proportion', fill='Spawn\nlocation')
```
**Qualitative observation**: the relative proportion of occurrences of eye states 
suggests that all messages have a common structure.

```{r}
contingency_table = t(table(messages))
contingency_table
```

$$
\begin{aligned}
  H_0&: \mbox{Aggregated counts of eye states are independent of the spawn location of the message} \\
  H_A&: \mbox{Aggregated counts of eye states depend on the spawn location of the message}
\end{aligned}
$$

```{r}
chisq.test(contingency_table)
```

The R output above reports a p-value of 0.91: fail to reject $H_0$, there is no
relation between location and relative proportions of eye states. This result supports
what was highlighted by the stacked bar chart earlier. It's then *reasonable* to 
think all messages share a common structure even if this structure is a random
variable in the probabilistic sense.

## Difference of sample proportions: eye mesasges

For any given message, the difference in relative proportions gives us information
about the prevalence of states, and more importantly an edge on the question 
*is my sample just a fluke?*

Pairwise difference of proportions,
$$
\begin{cases}
 p_0 - p_1,\, p_0 - p_2,\, p_0 - p_3,\, p_0 - p_4, \\
 p_1 - p_2,\, p_1 - p_3,\, p_1 - p_4,\, \\
 p_2 - p_3,\, p_2 - p_4,\, \\
 p_3 - p_4
\end{cases}
$$
```{r}
# This code is re-used in the following sections
# Run it to define the functions
suppressMessages(library(latex2exp))

compute_difference_of_sample_props = function(n, N, baseSample) {
  bsprops = array(dim=c(N, 10))
  for (i in 1:N) {
    props = as.vector(unlist(table(sample(baseSample, 
                                          n,
                                          replace=TRUE))/n, use.names=FALSE))
    assertthat::assert_that(sum(props) == 1)
    bsprops[i, 1]  = props[1] - props[2]
    bsprops[i, 2]  = props[1] - props[3]
    bsprops[i, 3]  = props[1] - props[4]
    bsprops[i, 4]  = props[1] - props[5]
    bsprops[i, 5]  = props[2] - props[3]
    bsprops[i, 6]  = props[2] - props[4]
    bsprops[i, 7]  = props[2] - props[5]
    bsprops[i, 8]  = props[3] - props[4]
    bsprops[i, 9]  = props[3] - props[5]
    bsprops[i, 10] = props[4] - props[5]
  }
  
  value = NULL
  statistic = NULL
  labels = c('p0 -p1', 'p0 - p2', 'p0 - p3', 'p0 - p4', 'p1 - p2', 
             'p1 - p3', 'p1 - p4', 'p2 - p3', 'p2 - p4', 'p3 - p4')
  for (i in 1:10) {
    value = c(value, bsprops[ , i])
    statistic = c(statistic, rep(labels[i], N))
  }
  
  list(plotdf=data.frame(value=value, statistic=statistic), 
       propdf=data.frame(bsprops))
}

suppressMessages(library(stringr))
suppressMessages(library(numform))

plot_and_CI95 = function(plotdf, propdf, sampleType) {
  p = ggplot(plotdf, aes(x=value, fill=statistic)) + 
        geom_histogram(binwidth=0.001, alpha=0.5) +
        scale_fill_brewer(palette='RdYlGn') +
        labs(title='Bootstrap statistic: difference of sample proportion',
             subtitle=str_interp('Sample: ${sampleType}'), 
             fill='Difference of\nsample proportion',
             x='Bootstrap statistic', y='Count') +
        # xlim(-0.1, 0.4) + 
        # ylim(0, 400) + 
        theme_minimal()
  
  CI = c(0.025, 0.975)
  for (i in 1:10) {
    interval = as.vector(unlist(quantile(propdf[[str_interp('X${i}')]], CI)))
    width = abs(interval[1] - interval[2])
    print(paste('Width:', f_pad_right(round(width, 3),pad.char='0', width=5),
                ' | CI 95 contains 0:', interval[1] <= 0 & interval[2] >= 0, ' | Mean:',
                round(mean(propdf[[str_interp('X${i}')]]), 4)))
  }
  
  return(p)
}
```

```{r}
res = compute_difference_of_sample_props(n=100000, 
                                         N=1000, 
                                         baseSample=messages$eye_state)
plot_and_CI95(res$plotdf, res$propdf, 'eye sample')
```

## Difference of sample proportions: actual text

Let's take a sample of text and run it by the same machinery. Refer to
the <code>data_wrangling</code> notebook for details regarding the choices made when 
converting to a 5-state system.

```{r}
res = compute_difference_of_sample_props(n=100000, 
                                         N=1000, 
                                         baseSample=strsplit(
                                           readLines('./encoded_text1.txt'),
                                           '')[[1]]
                                         )
plot_and_CI95(res$plotdf, res$propdf, 'actual text')
```

* ```encoded_text1.txt``` is English in a 5-state representation, 
* ```encoded_text2.txt``` is Latin encoded the same way. 
* ```encoded_text3.txt``` is a much larger block of English text encoded the same way

## Difference of sample proportions: uniformly distributed random numbers

```{r}
res = compute_difference_of_sample_props(n=100000, 
                                         N=1000, 
                                         baseSample=strsplit(
                                           readLines('./random_text.txt'),
                                           '')[[1]]
                                         )
plot_and_CI95(res$plotdf, res$propdf, 'random text')
```

## Difference of sample proportions: discord user Lymm's cipher

```{r}
res = compute_difference_of_sample_props(n=100000, 
                                         N=1000, 
                                         baseSample=strsplit(
                                           readLines('./lymm_cipher.txt'),
                                           '')[[1]]
                                         )
plot_and_CI95(res$plotdf, res$propdf, 'Lymm\'s cipher')
```